Modelling the emergence of whisker barrels

### 2020-01-28 paper/briefpaper/brief.tex
Last changes to text

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  The
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. \cmnt{The second term on the right
  quantifies the coupling between axon branching and synaptogenesis}. Here $\chi_i=0$ is a placeholder.

@@

\begin{figure}
\begin{fullwidth}
\includegraphics[width=\linewidth]{./MainFig_superlowres.png}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
  simulation approaching that obtained from barrels in \textbf{A} (dotted
  line); black squares measure the correspondence between the real and
  simulated barrel shapes, $\eta$ (the product of the sum of squared
  differences between real and simulated centers and the sum of differences in
  area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern
  in a Fgf8 misexpression experiment
  (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
  $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{fullwidth}
\end{figure}

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling fields. As
the number of distinct guidance fields is unlikely to approach the number of
barrels, further extensions were required.

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density \cmnt{of unconnected
  branches} at $t=0$
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}\,,
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. \cmnt{Note that this
  operation is local to individual afferent projections.}

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
center. As such, two orthogonal linear thalamic gradients were defined, from
which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41
barreloid centers recreated from \citealp{haidarliu_size_2001} (Fig.\,1B).

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time (Eq.\,\ref{eq:norm}).

The emergence of barrels in simulation required competition between thalamic
projections in terms of synaptic connectivity \cmnt{and an implicit self/other
  identifier amoungst projections}, and also in terms of cortical space (as
represented by $\chi$). This requirement may account for the absence of
barrels in rodents with larger brains, such as capybara, for which competition
for space is presumably weaker \citep{woolsey_comparative_1975}. Hence,
irrespective of whether barrels are necessary for adaptive whisker function,
the emergence of somatotopically ordered modular structures may be an
inevitable consequence of local competition for cortical territory driven by
input from an array of discrete sensory organs \citep{purves_iterated_1992}.

@@

The cortex was modelled as a two dimensional hexagonal lattice which
simplifies the computation of the 2D Laplacian. Within a boundary traced from
rat primary barrel cortex (Fig.\,1A) we set the hex-to-hex distance
$d$ to 0.03\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

@@

(i) The computation of the mean value of the Laplacian across one hexagon of
area $\Omega = \frac{\sqrt{3}}{2}d^2$, located at position $\mb{p}_0$, with
neighbours at positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \iint_\Omega \dvrg a_i(\mb{x}, t) dxdy & = \frac{1}{\Omega} \oint_c \frac{\partial a_i}{\partial \hat{\mb{n}}} d\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{mid} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big),
\end{split}
\end{equation}
%
where $d$ is the centre-to-centre distance between hexes in the grid, $v =
d/\sqrt{3}$ is the length of each of the 6 edges of the hexagon's perimeter
and $d\gamma$ is an infinitessimally small distance along the perimeter of the
hexagon.

@@

\begin{equation} \label{eq:divg}
\begin{split}
%
\frac{1}{\Omega} \iint_\Omega a_i\nabla\vcdot\mb{g}\;dxdy & = \frac{a_i(\mb{p}_0,t)}{\Omega}  \oint_c \mb{g}\vcdot d\hat{\mathbf{n}} \\
%
& \approx \frac{a_i(\mb{p}_0,t)}{\Omega} \sum_{j=1}^{6} \frac{\mb{g}_j + \mb{g}_0}{2}\vcdot \hat{\mb{n}}\;v \\
%
& = \frac{2 a_i(\mb{p}_0,t)v}{\sqrt{3}d^2} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
%%& = \frac{2 a_i(\mb{p}_0,t)d}{\sqrt{3}d^2\sqrt{3}} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
\Rightarrow \langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle & \approx \frac{a_i(\mb{p}_0,t)}{3d} \bigg( \sum_{j=1}^{6} \big(\mb{g}_j^x + \mb{g}_0^x\big) \cos \big(\frac{\pi}{3}(j-1)\big) + \sum_{j=1}^{6} \big({\mb{g}_j^y + \mb{g}_0^y}\big) \sin \big(\frac{\pi}{3}(j-1)\big) \bigg). \\
%
\end{split}
\end{equation}
%
Both this last expression, and the final expression of Eq.\,\ref{eq:lapl} are
easily computed, comprising short sums of nearest-neighbour relationships.

@@

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife_submission1}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James::compute\_divJ()} as \code{term1}, \code{term2} and \code{term3},
respectively.

>>>>

The barrel boundaries form a Voronoi tessellation across the cortical sheet
\citep{senft_mouse_1991} (Fig.\,1A), suggesting that barreloid topology is
preserved in the projection of thalamocortical axons into the cortex, and that
a barrel forms by lateral axon branching from an initial center-point that
ceases upon contact with axons branching from adjacent centers.  However, the
assumption of pre-arranged center-points is difficult to resolve with the
observation that axons arrive in the cortical plate as an undifferentiated
bundle, \emph{prior} to barreloid formation \citep{agmon_organized_1993}.

@@

\cite{karbowski_model_2004} developed a reaction-diffusion style model of how
extrinsic signalling gradients can constrain the emergence of distinct fields
from intrinsic cortical dynamics. Their model defines how the density of
connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a
1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by
$i$. The model was derived from the assumption that the rates at which $a_i$
and $c_i$ grow are reciprocally coupled. Extending the original 1D model to
simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and
$c_i(\mb{x},t)$, and model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. Note that computing the divergence in
simulation requires cells on the cortical sheet to communicate with immedately
adjacent cells only (see \emph{Materials \& Methods}) The second term on the
right quantifies the coupling between axon branching and synaptogenesis. Here
  $\chi_i=0$ is a placeholder.

@@

\begin{figure}
\begin{fullwidth}
\includegraphics[width=\linewidth]{./MainFig.png}
\caption{\textbf{A} Left shows a cytochrome oxidase stain obtained from rat S1
  by \cite{zheng_signal_2001}, with black lines to delineate barrels and to
  measure departure (Honda-$\delta$; see \citealp{senft_mouse_1991}) from a
  perfect Voronoi tesselation. Right shows the initial distribution of axon
  branching density ($a$) for one thalamocortical projection, and two
  molecular guidance fields ($\rho$). \textbf{B} The strengths of interaction
  $\gamma$ with fields $\rho_1$ and $\rho_2$ are indicated for each of 41
  projections by the lengths of green and blue arrows respectively, assuming
  that similar fields aligned to the posterior-anterior and medial-lateral
  axes in the ventroposterior medial nucleus of the thalamus are sampled at
  the locations of putative barreloid centers (reconstructed from
  \citealp{haidarliu_size_2001}). \textbf{C} Simulation results for parameters
  $N=41$, $\alpha=3$, $\beta=20$, $k=3$, $D=0.2$, $\gamma\in\pm 2$,
  $\epsilon=150$ and $\delta{t}=0.0001$. Colours indicate the thalamic
  projection for which the connection density is maximal, black lines
  delineate boundaries, and overlaid contours show $c>0.5$ (see Movie
  S1). \textbf{D} Red dots show the Honda-$\delta$ metric obtained from
  simulation approaching that obtained from barrels in \textbf{A} (dotted
  line); black squares measure the correspondence between the real and
  simulated barrel shapes, $\eta$ (the product of the sum of squared
  differences between real and simulated centers and the sum of differences in
  area; units mm$^4$). \textbf{E} Guidance fields and emergent barrel pattern
  in a Fgf8 misexpression experiment
  (c.f. \citealp{assimacopoulos_fibroblast_2012}), simulated by reflecting
  $\rho_1$ at the join ($\epsilon=80$). All scale bars 1mm.}
\label{fig:main}
\end{fullwidth}
\end{figure}

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling
fields. This is because localization of axon densities occurs only when
projections are influenced by interactions with two or more signalling
gradients that encourage migration in opposing directions. Note that these
dynamics are quite unline classic chemospecificity models
\citep{sperry_chemoaffinity_1963}, which essentially assume center-points
i.e., conditions in the target tissue that instruct pre-identified afferents
to stop growing. As the number of distinct guidance fields is unlikely to
approach the number of barrels, further extensions were required.

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial (random) density of unconnected
  branches at $t=0$
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}\,,
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. Note that this
  operation is local to individual afferent projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. As such, two orthogonal linear thalamic
gradients were defined, from which 41 pairs of $\gamma$ values were sampled,
at the coordinates of 41 barreloid centers recreated from
\citealp{haidarliu_size_2001} (Fig.\,1B).

@@

The present results suggest that the key requirements for the emergence of
realistic barrel patterning are i) at each cortical location thalamocortical
projections compete for a limited number of available synaptic connections
(Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each We speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
centerlocation the branching rate of a
given projection is reduced by the density of other projections
(Eq.\,\ref{eq:comp}), and iii) the branch density of each projection is
conserved over time (Eq.\,\ref{eq:norm}).

The emergence of barrels in simulation required cWe speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
centerompetition between thalamic
projections in terms of synaptic connectivity and an implicit self/other
identifier amoungst projections, and also competition in terms of cortical
space (as represented by $\chi$). This latter requirement may account for the
absence of
barrels in rodents with larger brains, such as capybara, for which competition
for space is presumably weaker \citep{woolsey_comparative_1975}. Hence,
irrespective of whether barrels are necessary for adaptive whisker function,
the emergence of somatotopically ordered modular structures may be an
inevitable consequence of local competition for cortical territory driven by
input from an array of discrete sensory organs \citep{purves_iterated_1992}.

@@

The cortical sheet was modelled as a two dimensional hexagonal lattice, which
simplifies the computation of the 2D Laplacian. Within a boundary traced
around the edge of a
rat barrel field (Fig.\,1A) we set the hex-to-hex distance
$d$ to 0.03\,mm, which resulted in a lattice containing 6515 hexes for the
simulations shown in Figs.\,1A,C \& D and 12739 hexes for the Fgf8
misexpression study shown in Fig.\,1E. Each hex contained 82 time-dependent
variables: 41 branching densities ($a_i$) and 41 connection densities ($c_i$).
The rate of change of each of the time-dependent variables (Eqs.\,\ref{eq:dc}
\& \ref{eq:da}) was computed using a fourth-order Runge-Kutta method.

@@

(i) The computation of the mean value of the Laplacian across one hexagon of
area $\Omega = \frac{\sqrt{3}}{2}d^2$, located at position $\mb{p}_0$, with
neighbours at positions $\mb{p}_1$--$\mb{p}_6$ is
%
\begin{equation} \label{eq:lapl}
\begin{split}
\langle D \dvrg a_i(\mb{p}_0,t) \rangle  = \frac{1}{\Omega} \oiint_\Omega \dvrg a_i(\mb{x}, t) \dif\Omega & = \frac{1}{\Omega} \oint \frac{\partial a_i}{\partial \hat{\mb{n}}} d\gamma \\
& \approx \frac{1}{\Omega} \sum_{j=1}^6 \frac{\partial a_i(\mb{p}_j)}{\partial \hat{\mb{n}}} \bigg\rvert_{mid} v \\
& = \frac{2}{\sqrt{3} d^2} \sum_{j=1}^6 \frac{a_i(\mb{p}_j) - a_i(\mb{p}_0)}{d} \frac{d}{\sqrt{3}} \\
& = \frac{2}{3 d^2} \sum_{j=1}^6 \big(a_i(\mb{p}_j) - a_i(\mb{p}_0)\big),
\end{split}
\end{equation}
%
where $v = d/\sqrt{3}$ is the length of each edge of the hexagon and $d\gamma$
is an infinitesimally small distance along its perimeter.

@@

\begin{equation} \label{eq:divg}
\begin{split}
%
\frac{1}{\Omega} \oiint_\Omega a_i\nabla\vcdot\mb{g}\;\dif\Omega & = \frac{a_i(\mb{p}_0,t)}{\Omega}  \oint \mb{g}\vcdot d\hat{\mathbf{n}} \\
%
& \approx \frac{a_i(\mb{p}_0,t)}{\Omega} \sum_{j=1}^{6} \frac{\mb{g}_j + \mb{g}_0}{2}\vcdot \hat{\mb{n}}\;v \\
%
& = \frac{2 a_i(\mb{p}_0,t)v}{\sqrt{3}d^2} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
%%& = \frac{2 a_i(\mb{p}_0,t)d}{\sqrt{3}d^2\sqrt{3}} \bigg( \sum_{j=1}^{6} \frac{\mb{g}_j^x + \mb{g}_0^x}{2} \vcdot  \hat{\mb{n}} + \sum_{j=1}^{6} \frac{\mb{g}_j^y + \mb{g}_0^y}{2} \vcdot  \hat{\mb{n}} \bigg) \\
%
\Rightarrow \langle a_i(\mb{p}_0,t)\nabla\vcdot\mb{g}(\mb{p}_0)\rangle & \approx \frac{a_i(\mb{p}_0,t)}{3d} \bigg( \sum_{j=1}^{6} \big(\mb{g}_j^x + \mb{g}_0^x\big) \cos \big(\frac{\pi}{3}(j-1)\big) + \sum_{j=1}^{6} \big({\mb{g}_j^y + \mb{g}_0^y}\big) \sin \big(\frac{\pi}{3}(j-1)\big) \bigg). \\
%
\end{split}
\end{equation}
%
Both this last expression, and the final expression of Eq.\,\ref{eq:lapl} can
be computed locally, by summing over values of the nearest neighbours.

@@

All code required to reproduce these results is available at
\url{https://github.com/ABRG-Models/BarrelEmerge/tree/eLife_submission1}. The
computations described in (i), (ii) and (iii) may be found in the class method
\code{RD\_James::compute\_divJ()} which calculates \code{term1}, \code{term2}
and \code{term3}, respectively.
